---
title: "R at Ecologic Institute"
author: "Albrecht Gradmann"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: ioslides_presentation
---

```{r setup,include=FALSE}
library(knitr)
library(png)
library(grid)
opts_chunk$set(message=FALSE,results='hide')
opts_knit$set(root.dir=normalizePath('../'))
```

```{r load_files,cache=FALSE,eval=T,echo=FALSE}
read_chunk('Example01/NUTS2choropletes_script.R')
read_chunk('Example02/RJSDMX_script.R')
```




## Introduction  
**R** is a free software environment for statistical computing and graphics. In this presentation I will show how to use R to create figures from Eurostat data in **SDMX** format. We will look at two examples:

* EU 2013 Unemployment rates by NUTS2 regions, percentage of total population displayed as a **choroplete map.**

* **Stacked bar-chart** of the structure of utilised agricultural area, 2007 (% of UAA).



## Using R with RStudio IDE
```{r whatisR, echo=FALSE}
img <- readPNG("resources/RStudio.png")
grid.raster(img)
```

## A large range of packages
In R several 1000 packages are available to facilitate all kinds of statistical analysis, data-work and visualizations.

```{r libraries,results='hide',tidy=FALSE,message=FALSE,warning=FALSE}
<<libraries>>
```


## SDMX a Standard for public data
Statistical Data and Metadata Exchange (2001)

```{r sdmx, echo=FALSE}
img <- readPNG("resources/sdmx.png")
grid.raster(img)
```



# Getting data
## Instead of the Eurostat Databrowser ...
```{r eurostat, echo=FALSE}
img <- readPNG("resources/Eurostat_REST.PNG")
grid.raster(img)
```

## ... use RJSDMX methods
This provides an entirely programmatic approach for data retrieval.
```{r get_data,message=FALSE,tidy=FALSE,cache=TRUE}
<<get_data>>
```

```{r convert_data,message=FALSE,tidy=FALSE,cache=TRUE}
<<convert_data>>
```

## Inspect the data-frame
```{r show_tsDf, results='markup',echo=2:3}
tsDF <- tsDf[c("FREQ","UNIT","SEX","GEO","TIME","OBS")] 
head(tsDF, n = 4)
str(tsDF, width=50, strict.width = "cut")
```

## Download geospatial data
Administrative boundaries as ESRI-shapefiles, a popular geospatial vector data format for geographic information system (GIS) software
```{r download_shp,message=FALSE,tidy=F,cache=TRUE}
<<download_shp>>
```

## Process geospatial data
Using contributed packages, Rs functionality can be extended to perform GIS functionalities.
```{r Read_shp,message=FALSE,tidy=F,cache=TRUE,warning=FALSE,results='hide'}
<<Read_shp>> 
```


## Merge and order data
As a final step in the data retrieval process, unemployment data has to be linked to the shapefiles.
```{r convert_shp,message=FALSE,tidy=F,warning=FALSE,results='hide',cache=TRUE}
<<convert_shp>>
```

```{r merge_data,message=FALSE,tidy=F,warning=FALSE,cache=TRUE}
<<merge_data>>
```


# Plotting
## Create a first draft
Once the data is in a suitable format, very few lines of code are needed to create an initial figure.
```{r first_plot,fig.show='hide',cache=TRUE}
<<first_plot>>
```

## Create a first draft
```{r first_plot2,echo=FALSE,fig.show='asis',cache=TRUE}
<<first_plot>>
```


## Put Europe into focus

```{r put_EU_in_focus,fig.show='hide',cache=TRUE}
<<put_EU_in_focus>>
```

## Put Europe into focus

```{r put_EU_in_focus2,echo=FALSE,fig.show='asis',cache=TRUE}
<<put_EU_in_focus>>
```

## Define a theme 
Themes allow to streamline appearance of graphs to consistently meet design guidelines
```{r window_dressing,message=FALSE,tidy=F, include=TRUE,echo=TRUE,eval=FALSE}
<<window_dressing>>
```

## Some final touches

```{r bin_data,fig.show='hide',include=FALSE,cache=TRUE}
<<bin_data>>  
```

* Bin data into classes (discrete scale)
* Choose colors 
* Re-position legend
* add country borders
* drop non-EU countries
* change projection

```{r set_colors,message=FALSE,tidy=F,include=FALSE}
<<set_colors>>
```

```{r add_borders,message=FALSE,tidy=F,include=FALSE}
<<add_borders>>
```

```{r change_projection,message=FALSE,tidy=F,include=FALSE}
<<change_projection>>
```

```{r window_dressing2,message=FALSE,tidy=F, include=TRUE,echo=FALSE}
<<window_dressing>>
```

```{r apply_theme,message=FALSE,tidy=F, include=TRUE,echo=FALSE}
<<window_dressing>>
```


## The final choroplete map

```{r lastTouch, include=FALSE}
<<lastTouch>>
```

```{r final_plot,echo=FALSE}
<<final_plot>> 
```

EU 2013 Unemployment rates by NUTS2 regions, percentage of total population. Source: [Eurostat (tgs00010)](https://github.com/Tungurahua/R_examples/blob/master/Example01/NUTS2choropletes_script.R)

## Facetted version of the map
```{r facetting,echo=FALSE}
<<facetting>> 
```

EU 2010-13 Unemployment rates by NUTS2 regions, percentage of total population Source: [Eurostat (tgs00010)](https://github.com/Tungurahua/R_examples/blob/master/Example01/NUTS2choropletes_script.R)

# Another example
## A figure published by Eurostat
```{r eurostatfig, echo=FALSE}
img <- readPNG("resources/fig25_eurostat.PNG")
grid.raster(img)
```

```{r getDatax,cache=TRUE,tidy=FALSE,include=FALSE}
<<getDatax>>
```

```{r getCodesVARIABLE,cache=TRUE,tidy=FALSE,echo=-1,include=FALSE}
<<getCodesVARIABLE>>
```

```{r headVARdic,results='markup',include=FALSE}
<<headVARdic>>
```

```{r dplyr, warning=FALSE,message=FALSE,tidy=FALSE,include=FALSE}
<<dplyr>>
```


## Reproduced in R {.smaller}

```{r firstplot2,fig.cap="An initial version of the plot.",fig.width=5,fig.height=5,warning=FALSE,tidy=FALSE,fig.margin=T,include=FALSE}
<<firstplotx>>
```

```{r theme2,message=FALSE,warning=FALSE,tidy=FALSE,include=FALSE}
<<themex>>
``` 
<div class="centered">
```{r windowdressing2,fig.width=5.5,fig.height=5.5, message=FALSE,warning=FALSE,tidy=FALSE,echo=FALSE}
<<windowdressingx>>
```
</div>


## Apply facets to improve readability {.smaller}
<div class="centered">
```{r betterplot,fig.width=5.5,fig.height=5.5,warning=FALSE,message=FALSE,echo=FALSE}
<<betterplot>>
```
</div>

## And move to web
```{r move_to_web,warning=FALSE,message=FALSE,echo=FALSE}
<<betterplot>>
```


`r I(plotly_iframe)`




# Concluding remarks

## Advantages of using R
* Reproducable work
* Unambiguous reference
* Re-using earlier work
* Workflow with early draft that is continuously refined
* Tight integration with [git](https://github.com/Tungurahua/R_examples)
* Drivers for different output formats (jpg, svg, etc) all the way down to interactive javascript visualizations with [rCharts](http://rcharts.io/) or [plot.ly](http://plot.ly)
* Work without altering source data

## Discussion
* Are Rs functionalities needed at Ecologic? 
* Would you use R yourself (proper training given)?
* Have you encountered R in project work?
* Do you see reproducibility requested in calls or projects?
* Which tools do you use to create figures, download data etc.? 
* ...





