---
title: "Creating NUTS-2 Choropletes from Eurostat Data"
author: "Albrecht Gradmann"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: rmarkdown::tufte_handout
---

```{r setup,include=FALSE}
library(knitr)
opts_chunk$set(message=FALSE,results='hide')
opts_knit$set(root.dir=normalizePath('../'))
```

```{r,cache=FALSE,eval=T,echo=FALSE}
read_chunk('Example01/NUTS2choropletes_script.R')
```


# Introduction
This document shows how to create a choroplete map from Eurostat Data using R. For this example I aimed to automate the entire process from downloading the data to the creation of the figure. The idea is to present an entirely reproducible process. The text shows the overall process from a birds-eye perspective. In particular it is not intended to serve as a general introduction to R[^quickR] or a discussion of the Pros and cons of using choropletes to display data. 

This tutorial shows the process of creating plot in R that visualizes the EU 2013 unemployment rates in the EU as percentage of total population. The final graphic will be a so-called choroplete map. This is a map where geographical regions are colored by some kind of metric. For this kind of figure to work well the individual regions should be roughly about the same size. For a more in-depth discussion you may refer to Leland Wilkinsons Grammar of Graphics [^Wilkinson] which we have available at Ecologic. The theoretical foundations presented in this book lay the foundations for the `ggplot2`package by Hadley Wickham [^Wickham], the approach used in this document 

I plan to create other handouts in the future. Please let me know if you would like to request a specific topic.

# Setting the scene
This tutorial gives a brief walk-through of the steps needed to create a choroplete map of EU 2013 unemployment rates by  NUTS-2 regions as a percentage of total population. Unemployment data as well as the geospatial data will be taken from Eurostat. Apart from unemployment data Eurostat provides several datasets on NUTS2 or even NUTS3 level. Adapting the present example to other datasets requires only minor modifications.

The example relies on a series of libraries which have to be installed on your system. Package installation in R is a simple and standard task and it can be done in several ways. For more information, you can either type  `help("install.packages")` in the R Console or look for one of the numerous tutorials on the internet[^howto_install]. Below is the series of commands you need to run first to attach the packages to your R session.

If you want to reproduce the results present in this document, you can copy-paste the code snippets into an R session. If you prefer the original source files, please send me an email and I will provide them happily. 

```{r libraries,results='hide',tidy=FALSE,message=FALSE}
<<libraries>>
```

```{r layout_libs,echo=FALSE,message=FALSE}
library(xtable)
```



# Getting the unemployment data
Eurostat provides several interfaces for access to data ranging from the online data-browser over the 'bulk-download-facility' to programmatic interfaces using `REST` or `SOAP`^[`REST` and `SOAP` the names of internet protocols. You don't have to bother with the details.]. While each of these interfaces has its own merits, a programmatic access to the `REST` interface is shown below. A programmatic access ensures that the creation of the figure is indeed reproducible. Provided access to the script used the process of creating the figure is unambiguously defined. In addition this approach allows the re-use and modularization of individual steps at a later stage.

We use the `getTimeSeries()` function to download the dataset from Eurostat. R provides some facilities to browse the Eurostat database.  The details of these facilities and an explanation how to define the arguments will be topic of another handout. The code below basically tells R to got to the Eurostat website, look for a dataset called `tgs00010` and from this to select annual data (that's the `A`), measured as percentage (the `PC` part) for both sexes (`T` stands for total) and for all available geographic entities (`*` means 'take all'). The `start` and `stop` arguments define the timespan for which we want to get the data.

```{r get_data,message=FALSE,tidy=FALSE,cache=TRUE}
<<get_data>>
```

The next step is to convert the downloaded data into a format suitable for further processing. Without going into too much detail, a `data-frame` is basically the way most R methods read data. It is basically a collection of variables of the same length.  
```{r convert_data,message=FALSE,tidy=FALSE,cache=TRUE}
<<convert_data>>
```

The table below shows the resulting data which is now stored in the `data-frame`. 

```{r show_tsDf, results='asis',echo=FALSE}
options(xtable.comment = FALSE)
options(xtable.booktabs = TRUE)
xtable(head(tsDf[c(2,3,5:8)]), caption = "First rows of tsDf")
```

If you ever browsed for data on the Eurostat website, you will probably recognize some of the variables:

\begin{itemize}
  \item \texttt{TIME} designates the year of the observation.
  \item \texttt{OBS} is the actual numeric observation value.
  \item \texttt{FREQ} tells us the observation frequency. Our data is annual (A) based on the data retrieval call.
  \item \texttt{UNIT}  hast been set to percentage (PC) in the data retrieval call.
  \item \texttt{SEX} is the total (T) for both sexes. \marginnote{Possible values for \texttt{SEX} are \underline{M}ale, \underline{F}emale and \underline{T}otal (not transgender as you might think).}
  \item \texttt{GEO} lists the geographic entities at NUTS-2 level.
\end{itemize}

Further inspection of the dataset reveals that figures for Turkey are included in the dataset. We use the following command to filter out all rows that start with TR. While in this case a relatively simple search string can be used, you should know that R is fully capable of interpreting regular expressions (REGEX)  allowing the definition of more sophisticated filter criteria. [^regex] the A last step we need to take for the 

```{r subset_data,message=FALSE,tidy=FALSE}
<<r subset_data>>
```

Now as we have gathered and prepared the unemployment data from Eurostat, our next step will be the acquisition of spatial data.

# Getting the empty maps
For the creation of choroplete maps we need information about the geographical extent of the NUTS-2 regions. This information is provided from Eurostat in the form of ESRI-shapefiles [^Eurostat_maps]. The appropriate shapefiles will be downloaded programmatically. As they are provided as zip-files we need to unpack them for further processing.



```{r download_shp,message=FALSE,tidy=F,cache=TRUE}
<<download_shp>>
```

Once the archive has been extracted, we can create the information into a `SpatialPolygonsDataframe` which is basically a processed shapefile ready for further treatment in R. We also set the map projection to WGS84. While a discussion of projections  [^projections]  is fay beyond the scope of this paper it is important to understand that this is a crucial step in any mapwork, especially if you aim to merge spatial data from different sources. Not knowing the projection of the data will then most likely lead to disaster.

```{r Read_shp,message=FALSE,tidy=F,cache=TRUE,warning=FALSE,results='hide'}
<<Read_shp>> 
```

As a last step we need to convert the `SpatialPolygonsDataframe` into a regular `dataframe`. This requirement comes from our usage of the `ggplot` package for doing the actual plotting routine, which expects `dataframes` as inputs.

```{r convert_shp,message=FALSE,tidy=F,warning=FALSE,results='hide'}
<<convert_shp>>
```

```{r show_eurMapDf, results='asis',echo=FALSE}
options(xtable.comment = FALSE)
options(xtable.booktabs = TRUE)
xtable(head(eurMapDf[]), caption = "First rows of tsDf",display = c("d","f","f","d","s","d","s","s"))
```

The resulting dataframe contains 7 Variables defining and naming the polygons for plotting:
\begin{itemize}
  \item long longitudinal coordinate of polygon edge.
  \item lat latitudinal coordinate of polygon edge.
  \item order order in which polygon edges are connected.
  \item hole  defines the type of polygon.
  \item piece defines part of polygon.
  \item group groups edges into polygon
  \item id lists the geographic entities at NUTS-2 level.
\end{itemize}

The first six variables contain the geometric definitions of the polygons. You will only probably never have to deal with them in any serious manner. The only exception is the variable `order` of which you have to make sure that it is sorted from smallest to largest before plotting. This step will be carried out in the upcoming section.

The last variable `id` however is crucial. It contains the names of all polygons contained in the shape file by their NUTS name. As you can see from the first lines displayed in the table they are not limited to NUTS-2 but include also all other levels of NUTS. We will use this variable to link the unemployment data to the geometry definition. This process is called merging and is explained in the nest section.

# Merge Unemployment Data with Map Data
Merging the unemployment and geographical data is achieved by telling R the respective datasets and name the columns to merge on. Note that this process of  join two data frames by one or more common key variables is what is called an 'inner join' in `SQL`. R also provides functions to carry out more complex joins. 

```{r merge_data,message=FALSE,tidy=F}
<<merge_data>>
```

As noted before, the records of the merged dataframe have to be sorted in the order in which they will be drawn to the graphics device.

```{r order_data,message=FALSE,tidy=F}
<<order_data>> 
```

With the merge of the two datasets we have concluded all steps in the process of gathering data and preparing it for plotting. 

# Plotting the data
The remainder of this document will deal with the process of a final plot from an initial draft. The steps shown provide examples of steps commonly carried out when producing a figure. It goes without saying that they are to a great extent matter of subjective taste. Maps are a special kind of display in which the general visualization is less strictly defined by the underlying data than e.g. a bar graph. While all of the following steps are in some way influenced by an underlying theory of perception, creating maps remains to a large extent an art. 

## Creating the first draft
Creating a first draft of the plot does not take much code. The syntax used is based on Leland Wilkinsons Grammar of Graphics as mentioned in the introduction. This may be a puzzling concept at first. In the end however it proves to be a tool that enables the user  to concisely describe the components of a graphic in a logical manner. 

```{r first_plot,fig.margin=T,fig.cap="An initial version of our map",}
<<first_plot>>
```

Although this first draft is  far from perfect it is already suitable to check if the graphic makes sense and actually reveals any interesting information. An initial inspection reveals that unemployment rates are highest in southern Spain and Greece. If this is a point we would like to make, we can further tune the graphic. While this is an initial version still, it can make sense to include it as a placeholder in a report draft. R can output the graphics in many different file formats [^graphicDev] that play well with different text editing and layout software. In Word you can link the figure data file so that it updates once a new version has been created in R.[^figLnk] 


## Putting Europe into focus
Our initial map contains regions which are far away from mainland Europe. These are mostly french DOM regions. To make the plot more readable we limit the graphic to mainland Europe.

```{r put_EU_in_focus,message=FALSE,tidy=F,fig.margin = TRUE}
<<put_EU_in_focus>>
```


If DOM or similar regions are of specific interest, they could be added as inset graphics at a later stage.


## Bin data into classes
R treats the unemployment figures as a continuous variable. To get a better feeling of the actual level of unemployment in the respective regions we can group the regions into different classes. This process is called \emph{binning} and is essentially a transformation into a discrete variable.


```{r bin_data,message=FALSE,tidy=F,fig.margin = TRUE}
<<bin_data>>  
```


If we re-plot the map with binned data we get an awkward looking result; especially when printed in the margin. By default the software assigns automatic values for the legend labels resulting in a crushed plot. 


## Fix color and legend
The default color palette is optimized for qualitative variables which have no intrinsic ordering. For our case this is not appropriate as the bins are ordered. For the selection of the appropriate color scale we turn to Cynthia Brewers website [^colorbrewer] which provides advice for choosing appropriate colors for mapping. The service suggests the palette 'OrRd' as being photocopy safe and appropriate for four classes of sequential data.


```{r set_colors,message=FALSE,tidy=F,fig.margin = TRUE}
<<set_colors>>
```


Re-coloring the classified data we can identify areas of high unemployment as red areas which actually makes sense as these are the areas that need attention.

The labeling text of the legend has also been adapted in this step. 

## Add country borders
Here we add the borders of regions an countries in different colors.


```{r add_borders,message=FALSE,tidy=F,fig.margin = TRUE}
<<add_borders>>
```


## Change projection
Maps can be displayed in different projections and they reveal a lot about the creators personality.[^xkcd_proj]


```{r change_projection,message=FALSE,tidy=F,fig.margin = TRUE}
<<change_projection>>
```


## Some window dressing
These final steps address the overall appearance of the figure. While there is a lot of fine tuning going on, it is possible to save this information to a \emph{theme} which can be applied to all figures in a given text. This allows to streamline graphics within a publication or to adhere to standards of corporate design.


```{r window_dressing,message=FALSE,tidy=F,fig.margin = TRUE}
<<window_dressing>>
```

# The Result
Below you will find the so-far final version of the plot. There are still several imperfections like a missing scale and a text box with source references.  Especially troublesome is the eye-catching hole in the Adriatic coast. While this could be due to rapid sink of the Kars Mountains followed by horrendous floods it is more likely an issue with the shape file addressing EU countries only.[^balkans] However all these issues are solvable and given that this is a tutorial and not the preparation of publication-ready graphics I will refrain from dealing with them here. 


```{r lastTouch, include=FALSE,cache=TRUE}
<<lastTouch>>
```




```{r final_plot, fig.cap = "EU 2013 Unemployment rates by NUTS2 regions, percentage of total population, both sexes Source: Eurostat (tgs00010)",echo=FALSE}
<<final_plot>> 
```



# Concluding remarks and Credits
One of the great advantages of R over other similar software packages is a very active and helpful user community. This document has been prepared by drawing on the following sources:


A cheatsheet for working with maps and ggplot [^ggmap_cheat]. This is a great concise resource that discusses how to include maps from Google, open street maps or stamen.

Stackoverflow[^SO] is a question and answer site for professional and enthusiast programmers. It's built and run by you as part of the Stack Exchange network of Q&A sites. Given that you post a well-defined question you will most likely get help here within hours.

A post on Max Marchis Blog[^MMarchi] was the source of the initial draft of this document. However this is just one of many. Searching for maps, R, choropletes and Eurostat will certainly reveal numerous other sources.

The `ggplot2`[^ggdoc] package has a great online documentation with many practical examples.


Last but not least this document has been created to get used to the Tufte-\LaTeX\ [^tufte_latex] document classes which define a style similar to the style Edward Tufte uses in his books and handouts. Tufte’s style is known for its extensive use of sidenotes, tight integration of graphics with text, and well-set typography. That said, if you need to print it, use color as it will be hard to read otherwise.


<!-- References --> 
[^Wickham]: Hadley Wickhams `ggplot2` is a very versatile and popular graphics package for R. http://ggplot2.org/
[^tufte_latex]: https://code.google.com/p/tufte-latex/

[^howto_install]: How to install packages in R http://tinyurl.com/nh7whtp
[^Eurostat_maps]: Administrative boundary shape files at Eurostat http://ec.europa.eu/eurostat/web/gisco/geodata/reference-data/administrative-units-statistical-units
[^SO]: Stack-Overflow - a community based help site. http://stackoverflow.com/
[^MMarchi]:Max Marchis Blog http://www.milanor.net/blog/?p=594
[^announce_eurostat]:http://rstudio-pubs-static.s3.amazonaws.com/27120_4dea44a84c9247c797289e145c17b38d.html
[^ggdoc]:Online Documentation of Hadley Wickhams `ggplot2` package. http://docs.ggplot2.org/current/
[^quickR]:One of many good introductory sites for R. http://www.statmethods.net/

[^Wilkinson]: Wilkinson: The Grammar of Graphics. 
Springer-Verlag New York, 2005.

[^rpubs_lovelace]:http://rpubs.com/RobinLovelace/ggmap
[^ggmap_cheat]:https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/ggmap/ggmapCheatsheet.pdf
[^colorbrewer]:Color advice for maps. http://colorbrewer2.org/
[^regex]: While intimidating at first regular expressions are a very powerful tool for any data work. http://www.regular-expressions.info/rlanguage.html
[^xkcd_proj]: xkcd explains what map projections tell about you http://xkcd.com/977/
[^projections]: An overview of map projections. http://www.colorado.edu/geography/gcraft/notes/mapproj/mapproj_f.html
[^dbjoins]: More information about database joins http://blog.codinghorror.com/a-visual-explanation-of-sql-joins/
[^graphicDev]: Output options from R http://www.cookbook-r.com/Graphs/Output_to_a_file/
[^figLnk]: How to link figures in Word. http://tinyurl.com/ccumxd
[^balkans]: Analysis of the availability and the quality of data on western balkans and Turkey - 2011 http://tinyurl.com/ngrhbpd
